Reconstructing contact network parameters from viral phylogenies

### 2016-06-25 doc/defence/presentation.tex
defences and papers and svm, oh my!

\usetikzlibrary{external}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.pathreplacing}

@@

\title{Phylogenetic inference of contact network parameters with kernel approximate Bayesian computation}
\author[RMM, RHL \& AFYP]{Rosemary M McCloskey$^1$ \and Richard H Liang$^1$ \and Art FY Poon$^{1,2}$}
\institute[UBC \& BCCfE]{$^1$BC Centre for Excellence in HIV/AIDS, Vancouver, Canada \\ $^2$Department of Medicine, University of British Columbia, Vancouver, Canada}
\date{HIV Dynamics \& Evolution, Woods Hole, USA, April 25, 2016}

@@

\begin{frame}{Most epidemiological models assume homogeneous mixing}
  \begin{columns}
    \begin{column}{0.6\textwidth}
      \centering
      \includegraphics[scale=0.6]{compartments}
      \vspace{1cm}

      \uncover<3->{\includegraphics[width=\textwidth]{sir-trajectories}}
    \end{column}
    \begin{column}{0.4\textwidth}
      \vspace{-2cm}
      \begin{itemize}
        \setlength{\itemsep}{24pt}
        \uncover<2->{
        \item Often provide a reasonable approximation in practice.
        }
        \uncover<3->{
        \item Can be inaccurate when substantial contact heterogeneity exists.
        }
      \end{itemize}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Network models capture contact heterogeneity}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth]{contactnet-empty}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
        \setlength{\itemsep}{12pt}
          \pause
        \item May offer more accurate predictions for highly structured
          populations.
          \pause
        \item Network parameters may be of interest for their own sake,
          \textit{e.g.} are there superspreaders?
          \pause
        \item Extremely difficult to estimate in practice.
      \end{itemize}
    \end{column}
  \end{columns}

\end{frame}

\begin{frame}{Contact networks shape transmission trees}
  \includegraphics[width=\textwidth]{contactnet}

  \begin{itemize}
      \pause
    \item Transmission trees in turn shape viral phylogenies.
      \pause
    \item \textbf{Aim:} estimate contact network parameters from viral
      phylogenies.
  \end{itemize}
\end{frame}

>>>>

\usetikzlibrary{external}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{calc}

@@

\title{Phylogenetic inference of contact network parameters with approximate Bayesian computation}
\author[RMM \& AFYP]{Rosemary M McCloskey$^1$ \and Bioinformatics Training Program \\ Supervisor: Art FY Poon$^{1,2}$}
\institute[UBC \& BCCfE]{$^1$BC Centre for Excellence in HIV/AIDS, Vancouver, Canada \\ $^2$Department of Medicine, University of British Columbia, Vancouver, Canada}
\date{M.Sc. Defence, July 26, 2016}

@@

\begin{frame}{Modeling in epidemiology}
  \centerline{\includegraphics[width=0.9\textwidth]{compartments}}
\end{frame}

\begin{frame}{Contact heterogeneity can affect epidemic expansion}
  \centerline{\includegraphics[width=0.9\textwidth]{sir-trajectories}}
\end{frame}

\begin{frame}{Contact networks capture contact heterogeneity}
  \begin{center}
  \begin{tikzpicture}
    \node (c) {\includegraphics[width=4cm]{contactnet-empty}};

    \uncover<2->{
    \node (a) [right=of c.north east, anchor=west] {nodes = individuals};
    \draw [->, thick, >=stealth] (c.north east) ++(-1.5cm, -0.5cm) -- (a.west);
    }

    \uncover<3->{
    \node (b) [below=0.5 of a] {edges = contacts};
    \draw [->, thick, >=stealth] (c.north east) ++(-0.75cm, -1.25cm) -- (b.west);
    }
  \end{tikzpicture}
  \end{center}
\end{frame}

\begin{frame}{Network models formally describe network structure}
  \begin{columns}
    \column{0.55\textwidth}
    \begin{minipage}[c][.6\textheight][c]{\linewidth}
      \begin{enumerate}
        \setlength{\itemsep}{12pt}
        \item Start with \only<1-4>{10}\only<5->{\alert{$n$}} women and 
          \only<1-4>{10}\only<5->{\alert{$n$}} men.
        \uncover<2->{
        \item Randomly form \only<2-4>{20}\only<5->{\alert{$p \times n$}} male-female partnerships.
        }
        \uncover<3->{
        \item Randomly rewire each edge with \only<3-4>{10}\only<5->{\\\alert{$q$ }}\% probability.
        }
      \end{enumerate}
    \end{minipage}
    \column{0.5\textwidth}
    \begin{minipage}[c][.6\textheight][c]{\linewidth}
      \only<1>{\includegraphics[width=5cm, page=1]{homophily}}
      \only<2>{\includegraphics[width=5cm, page=2]{homophily}}
      \only<3>{\includegraphics[width=5cm, page=3]{homophily}}
      \only<4->{\includegraphics[width=5cm, page=4]{homophily}}
    \end{minipage}
  \end{columns}

  \vspace{0.5cm}
  \begin{columns}
    \column{0.55\textwidth}
    \uncover<6->{\centerline{\large{network model}}}
    \column{0.55\textwidth}
    \uncover<6->{\centerline{\large{contact network}}}
  \end{columns}
\end{frame}

\begin{frame}{Barab\'asi-Albert model incorporates preferential attachment}
  \begin{columns}
  \column{0.5\textwidth}
  \begin{tikzpicture}[
      scale=2,
      every edge/.style={thick}
    ]
    \node (n1) at (1, 0) [circle, fill=black] { };
    \node (n2) at (0.86, 0.5) [circle, fill=black] { };
    \node (n3) at (0.5, 0.86) [circle, fill=black] { };
    \draw (n1) -- (n2) -- (n3) -- (n1);

    \uncover<2->{
      \node (n4) at (0, 1) [circle, fill=black] { };
    }
    \only<2-3>{
      \node [right=0 of n4.north east, anchor=south west] {$m = 2$};
    }
    \only<2-4>{
      \draw (n4) -- ++ (0.2, -0.2);
      \draw (n4) -- ++ (-0.2, -0.2);
    }

    \node [right=0 of n1] {%
      \only<3-4>{$\Pr \propto 2^\alpha$}%
      \only<6-7>{$\Pr \propto 3^\alpha$}%
    };
    \node [right=0 of n2] {%
      \only<3-4>{$\Pr \propto 2^\alpha$}%
      \only<6-7>{$\Pr \propto 3^\alpha$}%
    };
    \node [right=0 of n3] {%
      \only<3-4>{$\Pr \propto 2^\alpha$}%
      \only<6-7>{$\Pr \propto 2^\alpha$}%
    };

    \only<4>{
      \draw (n4) -- ++ (0.2, -0.2) -- (n2);
      \draw (n4) -- ++ (-0.2, -0.2) -- (n1);
    }

    \uncover<5->{
      \draw (n4) -- (n2);
      \draw (n4) -- (n1);
    }

    \uncover<6->{
      \node (n5) at (-0.5, 0.86) [circle, fill=black] { };
    }

    \only<6-7> {
      \draw (n5) -- ++ (0.2, -0.2);
      \draw (n5) -- ++ (-0.2, -0.2);
    }

    \uncover<6-7>{ \node [right=0 of n4.north east, anchor=south west] { $\Pr \propto 2^\alpha$ }; }

    \only<7>{
      \draw (n5) -- ++ (0.2, -0.2) -- (n3);
      \draw (n5) -- ++ (-0.2, -0.2) -- (n2);
    }

    \uncover<8->{
      \draw (n5) -- (n3);
      \draw (n5) -- (n2);
    }

    \uncover<9->{
      \node (n5) at (-0.86, 0.5) [circle, fill=black] { };
      \draw (n5) -- (n2);
      \draw (n5) -- (n1);
    }

    \uncover<10->{
      \node (n6) at (-1, 0) [circle, fill=black] { };
      \draw (n6) -- (n3);
      \draw (n6) -- (n4);
    }

    \uncover<11->{
      \node (n7) at (-0.86, -0.5) [circle, fill=black] { };
      \node (n8) at (-0.5, -0.86) [circle, fill=black] { };
      \node (n9) at (0, -1) [circle, fill=black] { };
      \node (n10) at (0.5, -0.86) [circle, fill=black] { };
      \node (n11) at (0.86, -0.5) [circle, fill=black] { };

      \draw (n7) -- (n2);
      \draw (n7) -- (n4);
      \draw (n8) -- (n2);
      \draw (n8) -- (n1);
      \draw (n9) -- (n2);
      \draw (n9) -- (n7);
      \draw (n10) -- (n3);
      \draw (n10) -- (n8);
      \draw (n11) -- (n2);
      \draw (n11) -- (n7);
    }
  \end{tikzpicture}

  \column{0.4\textwidth}
  \uncover<2->{
    $m$ = number of new edges per vertex \\
  }

  \uncover<3->{
    $\alpha$ = preferential attachment power \\
  }

  \uncover<11->{
    $N$ = number of nodes
  }
  \end{columns}
\end{frame}

\begin{frame}{Why fit network models?}
  \begin{columns}
    \column{0.33\textwidth}
    \includegraphics[width=0.85\textwidth]{sir-trajectories-vertical}

    \column{0.33\textwidth}
    \uncover<2->{
    \includegraphics[width=0.75\textwidth]{nettypes}
    }

    \column{0.33\textwidth}
    \uncover<3->{
    \begin{tikzpicture}
      \node (pic) { \includegraphics[width=0.75\textwidth]{vaccinate} };
      \node at (pic.east) { \includegraphics[width=1cm]{stock/syringe} };
    \end{tikzpicture}
    }
  \end{columns}

  \vspace{-0.5cm}

  \begin{columns}
    \column{0.33\textwidth}
    \uncover<1->{
    \centerline{\large predict}
    }

    \column{0.33\textwidth}
    \uncover<2->{
    \centerline{\large characterize}
    }

    \column{0.33\textwidth}
    \uncover<3->{
    \centerline{\large simulate}
    }
  \end{columns}
\end{frame}

\begin{frame}{Phylogenetic methods may help fit network models}
\end{frame}

\begin{frame}{Contact networks shape transmission trees}
  \begin{center}
    \begin{tikzpicture}
      [every node/.style = {inner sep=0pt},
       every path/.style = {->, >=stealth, very thick},
       label/.style = {anchor=south, color=white, inner sep=4pt},
       time/.style = {anchor=north west, inner sep=2pt}]
      \def\pht{1cm}
      \node (a) {\includegraphics[height=\pht]{stock/person}};
      \node at (a.south) [label] {\Large a};
      \node [below left=0.75 of a] (b) {\includegraphics[height=\pht]{stock/person}};
      \node at (b.south) [label] {\Large b};
      \node [below=0.75 of b] (e) {\includegraphics[height=\pht]{stock/person}};
      \node at (e.south) [label] {\Large e};
      \node [below right=0.75 of a] (c) {\includegraphics[height=\pht]{stock/person}};
      \node at (c.south) [label] {\Large c};
      \node [below=0.75 of c] (d) {\includegraphics[height=\pht]{stock/person}};
      \node at (d.south) [label] {\Large d};

      \fill [shade, opacity=0.2] (a.east) -- (c.north) -- (c.north west) -- (a.south east) -- cycle;
      \uncover<3->{
      \draw [blue] ($(a.east) + (0, -12pt)$) -- ($(c.north) + (-8pt, 0)$);
      }
      \fill [shade, opacity=0.2] (a.west) -- (b.north) -- (b.north east) -- (a.south west) -- cycle;
      \uncover<2->{
      \draw [red] ($(a.west) + (0, -12pt)$) -- ($(b.north) + (8pt, 0)$);
      }
      \fill [shade, opacity=0.2] (a.south) -- (a.south west) -- (e.north) -- (e.north east) -- cycle;
      \fill [shade, opacity=0.2] (b.south west) -- (b.south) -- (e.north) -- (e.north west) -- cycle;
      \fill [shade, opacity=0.2] ($(c.south) + (-8pt, 0)$) -- ++(16pt, 0) -- ($(d.north) + (8pt, 0)$) -- ++(-16pt, 0) -- cycle;
      \fill [shade, opacity=0.2] (b.east) -- (b.south east) -- (d.west) -- (d.north west) -- cycle;
      \uncover<4->{
      \draw [green!80!black] (c) -- (d);
      }

      \node (abcd) [right=6cm of a.north] {\includegraphics[height=\pht]{stock/person}};
      \node at (abcd.south) [label] {\Large a};
      \uncover<2->{
      \node (b) [below right=1 and 0.75 of abcd] {\includegraphics[height=\pht]{stock/person}};
      \node at (b.south) [label] {\Large b};
      \node (acd) [below left=1 and 0.5 of abcd] {\includegraphics[height=\pht]{stock/person}};
      \node at (acd.south) [label] {\Large a};
      }
      \uncover<3->{
      \node (cd) [below right=0.5 and 0.5 of acd] {\includegraphics[height=\pht]{stock/person}};
      \node at (cd.south) [label] {\Large c};

      \node (a) [below left=0.25 and 0.25 of acd] {\includegraphics[height=\pht]{stock/person}};
      \node at (a.south) [label] {\Large a};
      }
      \uncover<4->{
      \node (c) [below left=0.25 and 0.25 of cd] {\includegraphics[height=\pht]{stock/person}};
      \node at (c.south) [label] {\Large c};
      \node (d) [below right=0.4 and 0.4 of cd] {\includegraphics[height=\pht]{stock/person}};
      \node at (d.south) [label] {\Large d};
      }

      \uncover<2->{
      \draw [red] (abcd) -- (b);
      \draw [-, dashed] (abcd) -- (acd);
      }
      \uncover<3->{
      \draw [blue] (acd) -- (cd);
      \draw [-, dashed] (acd) -- (a);
      }
      \uncover<4->{
      \draw [-, dashed] (cd) -- (c);
      \draw [green!80!black] (cd) -- (d);
      }
    \end{tikzpicture}

  $\qquad$contact network \hfill transmission tree$\qquad$
  \end{center}
\end{frame}

\begin{frame}{Transmission trees shape viral phylogenies}
\end{frame}

\begin{frame}{Can we fit network models with viral phylogenies?}
\end{frame}

### 2016-06-25 doc/paper/abstract.tex
defences and papers and svm, oh my!

Models of the spread of disease in a population often make the simplifying
assumption that the population is homogeneously mixed, or is divided into
homogeneously mixed compartments. However, human populations have complex
structures formed by social contacts, which can have a significant influence on
the rate of epidemic spread. Contact network models capture this structure by
explicitly representing each contact which could possibly lead to a
transmission. We developed a method based on kernel approximate Bayesian
computation (kernel-ABC) for estimating structural parameters of the contact
network underlying an observed viral phylogeny. The method combines adaptive
sequential Monte Carlo for ABC, Gillespie simulation for propagating epidemics
though networks, and a kernel-based tree similarity score. We used the method
to fit the Barab\'{a}si-Albert network model to simulated transmission trees,
and also applied it to viral phylogenies estimated from five published HIV
sequence datasets. On simulated data, we found that the preferential attachment
power and the number of infected nodes in the network can often be accurately
estimated. On the other hand, the mean degree of the network, as well as the
total number of nodes, were not estimable with kernel-ABC. We observed
substantial heterogeneity in the parameter estimates on real datasets, with
point estimates for the preferential attachment power ranging from 0.2 to 1.1.
These results underscore the importance of considering contact structures when
performing phylodynamic inference. Our method offers the potential to
quantitatively investigate the network structure underlying viral epidemics,
using models appropriate to the context being studied.

>>>>

Models of the spread of disease in a population often make the simplifying
assumption that the population is homogeneously mixed, or is divided into
homogeneously mixed compartments. However, human populations have complex
structures formed by social contacts, which can have a significant influence on
the rate of epidemic spread. Contact network models capture this structure by
explicitly representing each contact which could possibly lead to a
transmission. We developed a method based on approximate Bayesian
computation (ABC) for estimating structural parameters of the contact network
underlying an observed viral phylogeny. The method combines adaptive sequential
Monte Carlo for ABC, Gillespie simulation for propagating epidemics though
networks, and a kernel-based tree similarity score. We used the method to fit
the Barab\'{a}si-Albert network model to simulated transmission trees, and also
applied it to viral phylogenies estimated from \del{five}\add{ten} published
HIV sequence datasets. On simulated data, we found that the preferential
attachment power and the number of infected nodes in the network can often be
accurately estimated. On the other hand, the mean degree of the network, as
well as the total number of nodes, were not estimable with ABC. We observed
substantial heterogeneity in the parameter estimates on real datasets, with
point estimates for the preferential attachment power ranging from 0.06 to
1.05. These results underscore the importance of considering contact structures
when performing phylodynamic inference. Our method offers the potential to
quantitatively investigate the contact network structure underlying viral
epidemics.

### 2016-06-25 doc/paper/coverletter.tex
defences and papers and svm, oh my!

\documentclass[12pt]{article}
\usepackage{fullpage}
\usepackage{garamondx}
\usepackage{parskip}
\usepackage{setspace}

% http://tex.stackexchange.com/questions/44618/dynamically-count-and-return-number-of-words-in-a-section
\newcommand\wordcount{\immediate\write18{texcount -sub=section \jobname.tex  | grep "Section" | sed -e 's/+.*//' | sed -n \thesection p > '/tmp/count.txt'}\ignorespaces\input{/tmp/count.txt}}

\title{Phylodynamic inference of contact network parameters through approximate Bayesian computation}
\author{Rosemary M. McCloskey, Richard H. Liang, and Art F.Y. Poon}

\doublespacing
\pagestyle{empty}
\begin{document}
\maketitle

\section{Abstract}

(\wordcount / 250 words)

Models of the spread of disease in a population often make the simplifying
assumption that the population is homogeneously mixed, or is divided into
homogeneously mixed compartments. However, human populations have complex
structures formed by social contacts, which can have a significant influence on
the rate of epidemic spread. Contact network models capture this structure by
explicitly representing each contact which could possibly lead to a
transmission. We developed a method based on kernel approximate Bayesian
computation (kernel-ABC) for estimating structural parameters of the contact
network underlying an observed viral phylogeny. The method combines adaptive
sequential Monte Carlo for ABC, Gillespie simulation for propagating epidemics
though networks, and a kernel-based tree similarity score. We used the method
to fit the Barab\'{a}si-Albert network model to simulated transmission trees,
and also applied it to viral phylogenies estimated from five published HIV
sequence datasets. On simulated data, we found that the preferential attachment
power and the number of infected nodes in the network can often be accurately
estimated. On the other hand, the mean degree of the network, as well as the
total number of nodes, were not estimable with kernel-ABC. We observed
substantial heterogeneity in the parameter estimates on real datasets, with
point estimates for the preferential attachment power ranging from 0.2 to 1.1.
These results underscore the importance of considering contact structures when
performing phylodynamic inference. Our method offers the potential to
quantitatively investigate the network structure underlying viral epidemics,
using models appropriate to the context being studied.

\section{Keywords}

Phylogenetics, phylodynamics, approximate Bayesian computation, contact network,
transmission tree, human immunodeficiency virus.

\section{Cover Letter}

Dear editors of \textit{Molecular Biology and Evolution},

On behalf of my coauthors and myself, I am submitting a manuscript entitled
``\textit{Phylodynamic inference of contact network parameters through
approximate Bayesian computation}'' for your consideration.

@@

All authors fulfill the authorship criteria, and have agreed to this
submission. \textit{Molecular Biology and Evolution} is our first choice for
this manuscript, which has not been submitted to any other journal.

Thank you for your time,

Rosemary M. McCloskey

\section{Reviewers}

\begin{enumerate}
  \item David Welch
  \item Gabriel Leventhal
  \item Erik Volz
  \item Tanja Stadler
  \item Oliver Ratmann
  \item Andrew Rambaut
\end{enumerate}

\section{Editors}

\begin{enumerate}
  \item Thomas Leitner
  \item Beth Shapiro
  \item Kieth Crandall
  \item John Novembre
\end{enumerate}

\section{Funding}

\begin{enumerate}
  \item Bill \& Melinda Gates Foundation (OPP1110049)
  \item Canadian Institutes for Health Research (HOP-111406)
\end{enumerate}

\section{Significant discoveries}

(\wordcount / 150 words)

Through a simulation study, we have found that certain parameters of the
contact network underlying a viral epidemic can be quantitatively estimated
from phylogenetic data. In particular, we investigated a preferential
attachment network model and were able to accurately infer the attachment power
and the number of infected nodes in the network, in most simulation scenarios
we considered. On the other hand, the average degree and total number of nodes
in the network had an insufficient effect on tree shape and could not be
quantitatively estimated. When applied five real-world HIV epidemics, we found
substantial heterogeneity of these parameters, which are known to have a
significant impact on the speed and pattern of epidemic spread.

\section{Significant methodological or theoretical advances}

(\wordcount / 150 words)

We have developed a phylodynamic method to quantitatively estimate the
parameters of contact network models from phylogenetic data. Our method is
based on approximate Bayesian computation, and combines adaptive sequential
Monte Carlo, Gillespie simulation, and a kernel-based tree similarity score.
Due to the use of the tree kernel, the method falls under the heading of
"kernel-ABC", which has been previously demonstrated to be a robust and
effective means of estimating epidemiological quantities from observed
phylogenies. To our knowledge, no other methods have been developed to estimate
network parameters from this type of data. Therefore, our method expands the
domain of phylodynamic inference to include network structure, offering
molecular epidemiologists a new perspective into ongoing epidemics.

\section{Significant new tools or resources}

(\wordcount / 150 words)

We provide a computer program "netabc" implementing our kernel-ABC method. The
C source code for the tool is freely available under a permissive license at
github.com/rmcclosk/netabc. This allows users to both apply the method as-is to
their data, and to incorporate new network models tailored to particular
research questions. The program is multithreaded, facilitating faster analysis
when more computing power is available. In addition, we provide two
supplemental programs to assist in phylodynamic analyses involving contact
networks: "nettree", which simulates a transmission tree over a contact network
by Gillespie simulation, and "treekernel", which quickly computes the kernel
similarity score between two trees. These tools will enable researchers to
quickly test hypotheses about the effect of network structure on transmission
tree shape.

\section{Broad impact}

(\wordcount / 150 words)

Phylodynamics and molecular epidemiology are relatively new areas of study, yet
they have undergone rapid development and have already seen public health
applications for real epidemics. Recent advances in sequencing technology have
allowed the collection of viral genetic data on a population scale in diverse
geographic and demographic settings. These data, together with phylodynamic
methods, provide an opportunity to gain insight into the real-world dynamics of
disease spread. Our work expands the domain of phylodynamic inference to
include structural parameters of the contact network underlying an epidemic.
Because our method is broadly applicable to any model from which networks can
be sampled, it will enable researchers to quantitatively investigate population
structure and contact heterogeneity in a variety of contexts.

\end{document}

>>>>

\documentclass[12pt]{letter}
\usepackage[T1]{fontenc}                                                         
\usepackage{garamondx}                                                           
\usepackage[garamondx,cmbraces]{newtxmath}

\signature{Rosemary M. McCloskey}
\address{BC Centre for Excellence in HIV/AIDS \\ 608-1081 Burrard Street \\ Vancouver, BC, Canada V6Z 1Y6}
\begin{document}
\begin{letter}{ }

\opening{Dear editors of \textit{Virus Evolution},}

On behalf of my coauthors and myself, I am submitting a manuscript entitled
``\textit{Reconstructing contact network parameters from viral phylogenies}''
for your consideration.

@@

\closing{Thank you for your time,}
\end{letter}
\end{document}
